---
title: "Getting started with ermeeth2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ermeeth2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ermeeth2)
```

`ermeeth2` gathers the functions used by the ThreeME team to run, post-process
and document ThreeME v4 simulations. It is the successor to `ermeeth`.

## The long-format data shape

Most of the plotting and table functions work on a long-format dataframe with
the columns:

| column | content |
|---|---|
| `year` | simulation year |
| `variable` | ThreeME variable code |
| `scenario` | scenario name |
| `values` | value of the variable in that scenario |
| `values_ref` | value of the variable in the baseline |
| `index_scen` | 1 for the baseline scenario, 0 otherwise |

`sector` and `commodity` are added when the variable code carries a sector or
commodity suffix.

## Where to start

- `long_data()` / `wide_data()` to reshape simulation output
- `threeme_transform()` for the ways a variable can be looked at: level,
  relative difference, absolute difference, percentage points, growth rate,
  index
- `simple_plot()` to draw a set of variables, over one or several scenarios
- `table_3me()` to build a result table, optionally comparing two versions of
  the model
- `run_simulations()` to run a set of scenarios

<!-- TODO: expand with a worked example on the sample data. -->
