---
title: "Graphs and tables with ermeeth2"
subtitle: "A tour of `simple_plot()` and `table_3me()`"
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 8
    fig-height: 4.5
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(ermeeth2)
library(dplyr)
```

## The data

Everything below runs on the small ThreeME v4 model shipped with the package, so
this document is reproducible without a simulation run. It is a genuine
`data_full` output: a baseline, one shock scenario `g`, five variables, 2016 to
2050.

```{r}
#| label: data
data_full <- readRDS(system.file("extdata", "minimodel.rds", package = "ermeeth2"))

glimpse(data_full)
```

The long format is what both functions expect: one row per
`year` × `variable` × `scenario`, with `values` for the scenario and
`values_ref` for the reference it is compared against. `sector` and `commodity`
are filled in when the variable code carries a suffix; here they are empty.

## Transformations

A ThreeME variable is rarely read raw. The ways of looking at one are declared
once, in `threeme_transformations()`, and both the plots and the tables draw
from that list.

```{r}
#| label: transformations
#| df-print: kable
purrr::map_dfr(threeme_transformations(), ~ tibble::as_tibble(.x))
```

`needs_ref` marks the transformations that compare against a reference
scenario; those also have `drop_baseline`, since the baseline compared to
itself is a row of zeroes. `percent` marks the ones held as a share and
displayed multiplied by 100.

You can apply them directly, but most of the time you will just pass the name
to `simple_plot()` or `table_3me()`.

```{r}
#| label: transform-direct
threeme_transform(data_full, "reldiff") |>
  filter(variable == "Y", year %in% 2021:2024) |>
  select(year, scenario, values, values_ref, value, unit)
```

## Graphs

### The default: relative difference to the baseline

One line per variable. The baseline is dropped automatically — it would be flat
at zero.

```{r}
#| label: fig-default
#| fig-cap: "GDP and its components, relative difference to the baseline"
simple_plot(data_full, c("Y", "CH", "I"))
```

### Naming the series

`labels` maps variable codes to something readable. Codes it does not mention
keep their code, so you can rename only what needs it. When the variable
dictionary exists, it will plug in here and this argument becomes the override
rather than the norm.

```{r}
#| label: fig-labels
#| fig-cap: "The same plot, with the series named"
simple_plot(
  data_full,
  variables = c("Y", "CH", "I"),
  labels = c(Y = "GDP", CH = "Household consumption", I = "Investment"),
  title = "Effect of the shock on activity"
)
```

### Other transformations

Growth rates, levels, an index based on a chosen year. Note that `level` keeps
the baseline, since there it is a series in its own right.

::: {.panel-tabset}

#### Growth rate

```{r}
#| label: fig-gr
simple_plot(data_full, "Y", transformation = "gr", labels = c(Y = "GDP"))
```

#### Level

```{r}
#| label: fig-level
simple_plot(data_full, "Y", transformation = "level", labels = c(Y = "GDP"))
```

#### Index, 2021 = 100

```{r}
#| label: fig-index
simple_plot(
  data_full, "Y",
  transformation = "index100", base_year = 2021,
  labels = c(Y = "GDP")
)
```

:::

With a single variable the colour goes to the scenario; with several it goes to
the variable and the scenario gets the linetype. `colour_by` overrides that if
you want the other way round.

### Several scenarios at once

The minimodel carries one shock, so here is a second one invented for the
demonstration — a milder version of the same shock.

```{r}
#| label: fake-scenario
milder <- data_full |>
  filter(scenario == "g") |>
  mutate(
    scenario = "g_mild",
    values = values_ref + (values - values_ref) / 2
  )

data_two <- bind_rows(data_full, milder)
```

```{r}
#| label: fig-scenarios
#| fig-cap: "One variable, two shocks"
simple_plot(
  data_two, "Y",
  labels = c(Y = "GDP"),
  title = "GDP under both shocks"
)
```

```{r}
#| label: fig-scenarios-multi
#| fig-cap: "Several variables and several scenarios: colour for the variable, linetype for the scenario"
simple_plot(
  data_two,
  variables = c("Y", "CH", "I"),
  labels = c(Y = "GDP", CH = "Consumption", I = "Investment")
)
```

### Colours

`palette` accepts a plain vector of colours, or a named one to pin particular
series. Names it does not cover fall back to the built-in palette.

```{r}
#| label: fig-palette
simple_plot(
  data_full,
  variables = c("Y", "CH", "I"),
  palette = c(Y = "#B61615", CH = "#017FC2", I = "#2E7437")
)
```

### Narrowing the window

```{r}
#| label: fig-window
simple_plot(data_full, c("Y", "CH"), startyear = 2021, endyear = 2035)
```

### Interactive

`interactive = TRUE` adds small hover points, each carrying the scenario, the
level under the shock, the level under the baseline, the growth rate and the
relative difference. The plot comes back through `ofce::girafy()`.

```{r}
#| label: fig-interactive
#| fig-cap: "Hover a point to read the underlying numbers"
simple_plot(
  data_full,
  variables = c("Y", "CH", "I"),
  labels = c(Y = "GDP", CH = "Consumption", I = "Investment"),
  interactive = TRUE
)
```

## Tables

### The default

Variables in rows, horizons in columns. `t` is the shock year, and the last
column is always the final year available.

```{r}
#| label: tbl-default
table_3me(data_full, c("Y", "CH", "I"))
```

The shock year is not passed: `table_3me()` finds it as the first year a
scenario departs from its reference.

```{r}
#| label: detect
detect_shock_year(data_full)
```

### Grouping the rows

Pass `variables` as a named list and the names become row groups, in the order
given.

```{r}
#| label: tbl-groups
table_3me(
  data_full,
  variables = list(
    Activity = c("Y", "CH", "I"),
    Policy = c("G", "MU")
  ),
  labels = c(Y = "GDP", CH = "Household consumption", I = "Investment",
             G = "Public spending", MU = "Mark-up")
)
```

### Mixing transformations in one table

The point of the unit column. Here activity is read in relative difference and
the mark-up in percentage points, in the same table.

```{r}
#| label: tbl-mixed
table_3me(
  data_full,
  variables = list(
    Activity = c("Y", "CH", "I"),
    Prices = c("MU")
  ),
  transformation = c(MU = "ppdiff", "reldiff"),
  labels = c(Y = "GDP", CH = "Household consumption", I = "Investment",
             MU = "Mark-up"),
  title = "Minimodel",
  subtitle = "Shock g, against the baseline"
)
```

The unnamed `"reldiff"` is the default for every variable the vector does not
name. The footnote is built from the transformations actually used.

### Choosing the horizons

```{r}
#| label: tbl-horizons
table_3me(
  data_full, c("Y", "CH", "I"),
  horizons = c(0, 5, 15, 25),
  labels = c(Y = "GDP", CH = "Consumption", I = "Investment")
)
```

Horizons past the end of the simulation are dropped rather than shown empty,
and the final year is appended as the long-term column.

### Several scenarios

Scenarios become column spanners.

```{r}
#| label: tbl-scenarios
table_3me(
  data_two, c("Y", "CH", "I"),
  labels = c(Y = "GDP", CH = "Consumption", I = "Investment"),
  horizons = c(0, 5, 10)
)
```

### Comparing two versions of the model

`data_secondary` takes a second `data_full`, typically the same scenarios run
after the equations changed. The two are laid out under a second spanner level.
Here the second database is again invented, as a model whose shock bites
slightly harder.

```{r}
#| label: fake-model
data_v2 <- data_full |>
  mutate(values = if_else(scenario == "baseline", values, values * 1.01))
```

```{r}
#| label: tbl-two-models
table_3me(
  data_full, c("Y", "CH", "I"),
  data_secondary = data_v2,
  model_names = c("v4.4", "v4.5"),
  labels = c(Y = "GDP", CH = "Consumption", I = "Investment"),
  horizons = c(0, 5, 10),
  caption = "The same shock under two versions of the model"
)
```

### Without the OFCE theme

`theme = "none"` returns a plain `gt` table, ready for your own styling.

```{r}
#| label: tbl-plain
table_3me(data_full, c("Y", "CH", "I"), theme = "none") |>
  gt::opt_stylize(style = 3, color = "gray")
```

## Clicking instead of typing

`threeme_viewer()` is an RStudio addin wrapping both functions. Point it at a
result file or at a `data_full` sitting in your session, tick the variables,
and it draws the plot and builds the table while exposing the arguments above
as controls — labels, per-variable transformations, groups, horizons, palette,
theme, and the second database for a two-model comparison.

```r
threeme_viewer()                                    # opens on its data panel
threeme_viewer(data_full)                           # starts from an object
threeme_viewer("data/output/minimodel.rds")         # starts from a file
```

It is also in the **Addins** menu, as *ThreeME viewer*.

The Code tab shows the two calls it has just run, so a session of clicking ends
with something to paste back into a quarto; inside RStudio the *Insert code*
button drops them at the cursor. The generated code is plain
`simple_plot()` / `table_3me()` — there is nothing in the viewer you cannot
write by hand.

## In short

```r
# a plot
simple_plot(data_full, c("Y", "CH", "I"))

# a table
table_3me(data_full, c("Y", "CH", "I"))
```

Everything else is optional. The arguments that matter most day to day:

| | `simple_plot()` | `table_3me()` |
|---|---|---|
| what to show | `variables`, `scenarios` | `variables` (list for groups), `scenarios` |
| how to read it | `transformation` (one) | `transformation` (one, or named by variable) |
| naming | `labels` | `labels` |
| period | `startyear`, `endyear` | `horizons`, `shock_year`, `end_year` |
| looks | `palette`, `colour_by`, `interactive` | `theme`, `digits`, `title`, `caption` |
| comparing models | — | `data_secondary`, `model_names` |
